Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches

### 2025-06-03 old.tex
revision complete and submitted

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
    \setmainfont[]{Helvetica}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% Break sections into new page
\usepackage{titlesec}
\newcommand{\sectionbreak}{\clearpage}
\usepackage{soul} % \hl{} = highlights

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning},
  pdfauthor={Anonymous1,; Anonymous1,2,},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning}
\author{Anonymous\textsuperscript{1,*} \and Anonymous\textsuperscript{1,2,*}}
\date{This version: June 03, 2025}

@@

\setstretch{1.5}
\textsuperscript{1} Anonymous Institute\\
\textsuperscript{2} Anonymous Institute

\textsuperscript{*} Correspondence: \href{mailto:Email}{Anonymous \textless{}Email\textgreater{}}, \href{mailto:Email}{Anonymous \textless{}Email\textgreater{}}

@@

\begin{itemize}
\item
  An experiment was conducted where K-12 students (N=122) used AI chatbots to solve school-related problems.
\item
  Two interventions were tested: (1) comparing AI-generated step-by-step explanations to predictions alone, and (2) contrasting AI chatbot support inspired by the Socratic method (guided questioning) versus non-Socratic (direct answers) support.
\end{itemize}

\begin{itemize}
\item
  Participants performed better under step-by-step explanations. They also engaged more frequently with the Socratic AI, though this approach did not result in improved performance and retain of knowledge compared to the non-Socratic AI.
\item
  While students found interactions with the AI useful, they perceived the Socratic AI as less helpful than non-Socratic overall.
\item
  The findings highlight challenges in designing AI tutors that effectively foster critical thinking while maintaining student satisfaction, raising concerns about their adoption in educational settings.
\end{itemize}

@@

How does integrating large language models (LLMs) into classroom activities influence students' learning? To investigate this question, we conducted a randomized experiment with students aged 14-16 (N=122), testing two interventions: (i) comparing AI-generated solutions with step-by-step explanations versus solutions alone and (ii) Socratic AI chatbots promoting critical thinking through incremental guidance versus non-Socratic AI offering direct solutions. Students completed school-related tasks under these conditions. We collected students' performance and interaction logs data, as well as their attitudes through self-reported surveys. Results indicated that AI-generated explanations significantly improved students' performance over solutions alone, highlighting the benefits of step-by-step guidance. The Socratic AI approach fostered significantly greater engagement and interaction. However, it did not achieve significant improvements in learning, and a higher fraction of students perceived it as less helpful. Furthermore, despite overall positive perceptions of AI assistance, students exhibited limited retention, failing to apply learned concepts to new situations without the aid of AI. These findings underscore both the potential and limitations of LLM-based chatbots in K-12 education, especially in balancing the development of critical thinking with user satisfaction.

\textbf{Keywords}: artificial intelligence; large language models; learning; education policy; experiment; k-12

@@

Integrating artificial intelligence (AI) into educational settings has been debated for over three decades (\citeproc{ref-roll2016evolution}{Roll and Wylie 2016}). Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices (\citeproc{ref-yan2024promises}{Yan, Greiff, et al. 2024}; \citeproc{ref-kasneci2023chatgpt}{Kasneci et al. 2023}; \citeproc{ref-tlili2023if}{Tlili et al. 2023}; \citeproc{ref-debets2025chatbots}{Debets et al. 2025}). Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of academic tasks: facilitating problem-solving (\citeproc{ref-urban2024chatgpt}{Urban et al. 2024}), providing personalised assistance (\citeproc{ref-darvishi2024impact}{Darvishi et al. 2024}), language learning (\citeproc{ref-derakhshan2024chatgpt}{Derakhshan and Ghiasvand 2024}; \citeproc{ref-song2023enhancing}{Song and Song 2023}), and evaluating students' work (\citeproc{ref-henkel2024can}{Henkel et al. 2024}). However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks (\citeproc{ref-farrokhnia2024swot}{Farrokhnia et al. 2024}; \citeproc{ref-weidinger2022taxonomy}{Weidinger et al. 2022}; \citeproc{ref-walczak2023challenges}{Walczak and Cellary 2023}).

A key advancement over previous AI systems is that LLMs can provide step-by-step reasoning to answer students' queries (\citeproc{ref-wei2022chain}{Wei et al. 2022}) and that by adding ``Let's think step by step'\,`, LLMs can improve their performance significantly (\citeproc{ref-kojima2022large}{Kojima et al. 2022}). Although evidence suggests that such explanations can foster children's learning and contribute to the development of their scientific reasoning (\citeproc{ref-dejong1986explanation}{DeJong and Mooney 1986}; \citeproc{ref-legare2014contributions}{Legare 2014}; \citeproc{ref-danovitch2021mind}{Danovitch et al. 2021}), there is limited empirical research examining how such AI-generated explanations impact students' learning processes. Most prior studies have not isolated the specific role of detailed AI-generated reasoning in enhancing problem-solving skills. To fill this gap, we conducted a randomised experiment with k-12 students (n = 122) who were engaged in a simple estimation task -- guessing the value of coins in a jar. Half of the participants received only an AI-generated prediction, while the other half were additionally provided with AI-generated explanations detailing how to derive such estimation. This setup enabled to address our first research question (RQ):

\textbf{RQ1: Do AI-generated explanations enhance students' problem-solving performance in school tasks, specifically in terms of estimation accuracy?}

The distinction between AI-generated \emph{explanations} and \emph{solutions} further allows us to evaluate whether providing explanations influences students' trust and judgment of AI-generated content. Indeed, LLMs can generate inaccurate solutions and flawed reasoning (\citeproc{ref-saxton2019analysing}{Saxton et al. 2019}; \citeproc{ref-kalyan2021how}{Kalyan et al. 2021}; \citeproc{ref-hendrycks2021measuring}{Hendrycks et al. 2021}). However, there is evidence that providing explanations increases users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear (\citeproc{ref-rai2020explainable}{Rai 2020}; \citeproc{ref-ferrario2022explainability}{Ferrario and Loi 2022}). In an educational setting, this may affect learning outcomes in a complex manner. Students may interpret well-written explanations as evidence that an answer is correct, even when it is not. Conversely, they may notice minor errors or inconsistencies in an explanation that reduce their trust, even if the overall solution is accurate. This situation motivates our second research question:

\textbf{RQ2: How do AI-generated explanations influence students' perceived credibility of AI-generated solutions?}

Another critical technical progress of contemporary LLMs over previous AI systems is their ability to follow specific instructions when engaged in conversations. Building on this capability, another goal of this study is to examine the impact of two fundamental modes of interaction on students' learning outcomes: one that fosters critical thinking through incremental guidance (``Socratic AI'') and another that offers immediate solutions (``non-Socratic AI'').

Although classical pedagogical methods, such as the Socratic Method, have long been valued for promoting critical thinking and a deeper understanding (\citeproc{ref-lam2011socratic}{Lam 2011}; \citeproc{ref-dalim2022promoting}{Dalim, Ishak, and Hamzah 2022}; \citeproc{ref-wilberding2021socratic}{Wilberding 2021}; \citeproc{ref-cleveland2015beyond}{Cleveland 2015}), there is a notable gap in the AI education literature regarding how to integrate these methods into AI-driven solutions. Existing AI tools, such as ChatGPT, often focus on providing direct solutions or feedback, which can potentially limit opportunities for active student engagement and reflection. Our study addresses this gap by conducting a parallel experimental intervention that compares Socratic-style AI interactions with more conventional, solution-focused AI interactions in a K-12 setting. This comparison contributes novel evidence on how different AI interaction modes can influence student learning processes and outcomes, informing the design of more pedagogically sound AI educational systems.

More specifically, building on previous literature suggesting that the Socratic method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning (\citeproc{ref-cleveland2015beyond}{Cleveland 2015}), we test the following research questions:

\textbf{RQ3: Do student-AI interactions guided by the Socratic method promote better performance?}

\textbf{RQ4: Do student-AI interactions guided by the Socratic method enhance students' confidence in their answers?}

In addition to educational outcomes, exploring the trade-offs between Socratic and non-Socratic AI raises an important question about student \emph{satisfaction}. If students find interacting with Socratic AI chatbots unengaging or frustrating, such perceptions might drive students away from Socratic AI to seek other (non-Socratic) tools. Indeed, research on conversation length with LLM-powered chatbots has shown that people tend to have mixed reactions when chatbots are instructed to engage in more extended conversations (\citeproc{ref-huang2024does}{Huang et al. 2024}). Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools. This consideration leads to our last research question:

\textbf{RQ5: How do students perceive the helpfulness of Socratic AI compared to non-Socratic AI?}

Addressing these questions aims to contribute to a growing debate on using pedagogical principles and techniques, like the Socratic approach, to integrate AI in education, which has recently gained considerable attention. For instance, Khan Academy has developed its AI tutor, Khanmigo, based on Socratic principles, aiming to foster critical thinking and engagement in learners.\footnote{\url{https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor}} Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny (\citeproc{ref-lara2020artificial}{Lara and Deckers 2020}).\footnote{See also \url{https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge}} Beyond education, Socratic AI systems are also being developed to foster critical thinking among the general public, particularly in efforts to combat disinformation (\citeproc{ref-duelen2024socratic}{Duelen, Jennes, and Van den Broeck 2024}).

\subsection{Literature Review}\label{literature-review}

A growing body of research highlights the tremendous potential of LLMs to improve student learning outcomes. A recent scoping review identifies 53 distinct use cases for LLMs in education (\citeproc{ref-yan2024practical}{Yan, Sha, et al. 2024}), ranging from automatic grading and personalised feedback to teaching support and content generation, reflecting the high versatility and scalability of LLMs. Within this expanding literature, this study centres on \emph{AI chatbots}, conversational agents powered by LLMs, and their educational applications, reflecting their increasing prominence and potential impact in this field.

Early investigations into the use of AI chatbots, such as ChatGPT, in educational settings have identified both opportunities and challenges (\citeproc{ref-farrokhnia2024swot}{Farrokhnia et al. 2024}; \citeproc{ref-noroozi2024generative}{Noroozi et al. 2024}). On the one hand, ChatGPT can serve as a personalised tutor, assist in drafting assignments, or generate creative prompts for classroom activities. On the other hand, concerns have been raised around academic integrity, overreliance on AI-generated content, and the need for critical thinking skills when interpreting chatbot outputs.

A recent review of over 70 studies provides a comprehensive overview of how chatbots are used in various educational contexts (\citeproc{ref-debets2025chatbots}{Debets et al. 2025}). It classifies chatbot applications by roles, including tutoring, giving feedback, and helping with administrative tasks, and examines how they are designed and utilised. The review finds that AI chatbots generally meet their goals. For example, a quasi-experimental study with 320 middle schoolers found that an AI teaching chatbot significantly improved math knowledge (\citeproc{ref-xing2025development}{Xing et al. 2025}). Similarly, a meta-analysis of 24 RCTs shows that AI chatbots significantly enhance student outcomes, with stronger effects in higher education (\citeproc{ref-wu2024ai}{R. Wu and Yu 2024}). However, most studies are short-term, may be subject to publication bias, and often lack a strong theoretical foundation.

Despite promising results, key limitations remain. For instance, Er et al. (\citeproc{ref-er2024assessing}{2024}) found that students in a programming class performed better with instructor feedback, which was seen as more useful and fair than AI-generated feedback, highlighting the current gap between AI and human instruction. Similarly, evidence of performance boosts from non-education settings is mixed. A meta-analysis by Vaccaro, Almaatouq, and Malone (\citeproc{ref-vaccaro2024combinations}{2024}) found that human-AI teams often performed worse than either humans or AI alone, suggesting a more nuanced relationship. In education, additional concerns are that integrating AI might encourage cheating, over-reliance, or superficial understanding (\citeproc{ref-farrokhnia2024swot}{Farrokhnia et al. 2024}; \citeproc{ref-yusuf2024generative}{Yusuf, Pervin, and Román-González 2024}; \citeproc{ref-eke2023chatgpt}{Eke 2023}; \citeproc{ref-lee2024cheating}{Lee et al. 2024}), a risk that is difficult to evaluate without longitudinal studies.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
    \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% Break sections into new page
\usepackage{titlesec}
\newcommand{\sectionbreak}{\clearpage}

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches},
  pdfauthor={Andrea Blasco1,; Vicky Charisi1,2,},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches}
\author{Andrea Blasco\textsuperscript{1,*} \and Vicky Charisi\textsuperscript{1,2,*}}
\date{This version: June 03, 2025}

@@

\setstretch{2}
\textsuperscript{1} European Commission, Joint Research Centre, Brussels, Belgium\\
\textsuperscript{2} Harvard University, Berkman Klein Center, Cambridge, MA, USA

\textsuperscript{*} Correspondence: \href{mailto:andrea.blasco@ec.europa.eu}{Andrea Blasco \textless{}\href{mailto:andrea.blasco@ec.europa.eu}{\nolinkurl{andrea.blasco@ec.europa.eu}}\textgreater{}}, \href{mailto:vcharisi@law.harvard.edu}{Vicky Charisi \textless{}\href{mailto:vcharisi@law.harvard.edu}{\nolinkurl{vcharisi@law.harvard.edu}}\textgreater{}}

@@

\begin{itemize}
\item
  An experiment was conducted where K-12 students used AI chatbots to solve school-related problems.
\item
  Two AI chatbot approaches were tested: one provided incremental guidance to encourage critical thinking (``Socratic''), while the other offered immediate solutions (``non-Socratic'').
\item
  Results showed that AI-generated explanations improved students' performance over solutions without them, highlighting the value of AI-generated guidance.
\item
  Participants engaged more frequently with the Socratic AI, though this did not result in improved performance compared to the non-Socratic AI.
\item
  While students found interactions with the AI useful, they perceived the Socratic AI as less helpful overall.
\item
  The findings highlight challenges in designing AI tutors that effectively foster critical thinking while maintaining student satisfaction, raising concerns about their adoption in educational settings.
\end{itemize}

@@

How does integrating large language models (LLMs) into classroom activities influence students' learning? To address this question, we conducted a randomized experiment comparing two distinct chatbot approaches: one designed to encourage critical thinking through incremental guidance (``Socratic AI'') and another providing immediate solutions (``non-Socratic AI''). The study involved students aged 14 to 16, who engaged in various school-related tasks under these experimental conditions. Students' attitudes were measured through self-reported surveys. Results indicated that AI-generated explanations significantly improved students' performance over solutions provided without such explanations, highlighting the benefits of step-by-step guidance. The Socratic AI approach fostered significantly greater engagement and interaction. However, it did not achieve significant improvements in learning, and a higher fraction of students perceived it as less helpful. Furthermore, despite students generally perceiving AI assistance as beneficial, they exhibited limited retention, failing to apply learned concepts to new situations when we removed AI assistance. These findings contribute to the ongoing debate on integrating LLM-powered chatbots in education, highlighting key challenges in designing AI tutors that effectively foster critical thinking while maintaining student satisfaction, raising concerns about their adoption in educational settings.

\textbf{Keywords}: artificial intelligence; large language models; learning; education policy; experiment, k-12

@@

Integrating artificial intelligence (AI) into educational settings has been debated for over three decades (\citeproc{ref-roll2016evolution}{Roll and Wylie 2016}). Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices (\citeproc{ref-yan2024promises}{Yan, Greiff, et al. 2024}; \citeproc{ref-kasneci2023chatgpt}{Kasneci et al. 2023}; \citeproc{ref-tlili2023if}{Tlili et al. 2023}). Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of educational tasks: facilitating problem-solving (\citeproc{ref-urban2024chatgpt}{Urban et al. 2024}), language learning (\citeproc{ref-derakhshan2024chatgpt}{Derakhshan and Ghiasvand 2024}; \citeproc{ref-song2023enhancing}{Song and Song 2023}), or evaluation of students' work (\citeproc{ref-henkel2024can}{Henkel et al. 2024}). However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks. This study addresses two critical aspects of AI integration in education: the effectiveness of AI-generated explanations and the optimal modes of interaction between AI and students, specifically whether Socratic or alternative approaches yield better outcomes.

A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries. This study investigates how this ability influences students' interactions and helps them enhance their problem-solving skills. Specifically, we first examine the impact of AI explanations on students' performance in a numerical estimation task and their assessment of the accuracy of AI-generated predictions. Then, we test two approaches to access these explanations through AI chatbots: one approach that encourages critical thinking through incremental guidance (``Socratic AI'') and another providing immediate solutions (``non-Socratic AI'').

Previous research has shown that explanations can foster children's causal learning and contribute to the development of their scientific reasoning (\citeproc{ref-dejong1986explanation}{DeJong and Mooney 1986}; \citeproc{ref-legare2014contributions}{Legare 2014}; \citeproc{ref-danovitch2021mind}{Danovitch et al. 2021}). Similarly, explanations generated by LLMs may guide students in their problem-solving, for example, by decomposing a numerical estimation task into more manageable steps, such as considering heuristics or approximations. However, LLMs can also generate inaccuracies and flawed reasoning (\citeproc{ref-saxton2019analysing}{Saxton et al. 2019}; \citeproc{ref-kalyan2021how}{Kalyan et al. 2021}; \citeproc{ref-hendrycks2021measuring}{Hendrycks et al. 2021}), misleading students if followed without scrutiny. Understanding this dual role of AI explanations as a valuable educational tool and a potential source of misinformation leads to our first research question (RQ):

\textbf{RQ-1: Do AI-generated explanations enhance learning by providing insights, or do they risk impairing students' critical thinking by presenting inaccurate reasoning as authoritative?}

The study also aims to explore how different modes of student-AI interaction more broadly influence students' performance and perceptions, of which providing explanations is one key element. Specifically, we contrast two fundamental approaches: the ``Socratic'' and ``non-Socratic'' methods. The Socratic approach, inspired by the Socratic teaching method (\citeproc{ref-lam2011socratic}{Lam 2011}; \citeproc{ref-dalim2022promoting}{Dalim, Ishak, and Hamzah 2022}), engages students through an argumentative dialogue where the AI tutor asks thought-provoking questions to stimulate critical thinking and self-reflection (\citeproc{ref-wilberding2021socratic}{Wilberding 2021}). This method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning (\citeproc{ref-cleveland2015beyond}{Cleveland 2015}). In contrast, the non-Socratic approach directly answers students' queries without necessarily engaging them in a dialogue or exploration.

The Socratic approach has recently gained considerable attention in the context of AI tutoring in education. Notably, Khan Academy has developed its AI tutor, Khanmigo, based on these principles, aiming to foster critical thinking and engagement in learners.\footnote{\url{https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor}}. Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny (\citeproc{ref-lara2020artificial}{Lara and Deckers 2020}).\footnote{See also \url{https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge}} This debate raises the following research question:

\textbf{RQ-2: Do student-AI interactions guided by the Socratic method promote deeper critical thinking and enhance students' confidence in their answers?}

Another objective of comparing Socratic and non-Socratic AI is to assess the extent of ``user satisfaction'' among students. Regardless of its academic benefits, Socratic AI could find limited application if students find interacting with Socratic AI chatbots unengaging or frustrating. Negative interactions or perceptions might drive students away in favour of more straightforward (non-Socratic) tools that provide direct answers, such as ChatGPT. Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools. This leads to our last research question:

\textbf{RQ-3: How do students perceive the helpfulness of Socratic AI compared to non-Socratic AI?}

By focusing on the impact of AI explanations and the Socratic AI approach on students' critical thinking, we aim to contribute to a growing literature about integrating AI into education. Recent research on AI chatbots and LLMs indicates significant potential for improving student learning outcomes. A meta-analysis of 24 randomised studies highlights the positive impact of chatbots, especially on student productivity and learning engagement (\citeproc{ref-wu2024ai}{Wu and Yu 2024}). However, several studies are more cautious about the benefits, partly due to concerns that AI might encourage cheating, over-reliance, or superficial understanding, which are challenging to measure in short-term studies (\citeproc{ref-lee2024cheating}{Lee et al. 2024}). Additional studies on LLMs' impact on academic integrity show that students and teachers struggle to differentiate between human and AI-generated text, raising risks of increased academic dishonesty and eroded trust (\citeproc{ref-yusuf2024generative}{Yusuf, Pervin, and Román-González 2024}; \citeproc{ref-eke2023chatgpt}{Eke 2023}). Furthermore, as LLMs continue to evolve, students and educators may lack the skills to leverage them effectively, which could result in missed educational opportunities (\citeproc{ref-kaplan2023generative}{Kaplan-Rakowski et al. 2023}).

The ethical and societal implications of LLMs in education are also complex. Issues such as bias, privacy, surveillance, and autonomy persist, especially in K-12 contexts, where the impact of algorithmic decisions on young learners' agency and privacy can be profound (\citeproc{ref-williams2024ethical}{Williams 2024}). LLMs introduce unique challenges, notably in alignment, where the system's goals may not align fully with educational objectives, despite companies' efforts to mitigate such risks through fine-tuning and reinforcement learning (\citeproc{ref-gabriel2020artificial}{Gabriel 2020}). Concerns also extend to students' digital well-being, as AI in education could inadvertently contribute to unhealthy digital habits and reliance on AI assistance over critical thinking (\citeproc{ref-weidinger2022taxonomy}{Weidinger et al. 2022}). Given these multidimensional impacts, further research is needed to optimise AI's role in education, balancing short-term learning gains with long-term skill development and ethical considerations.

Research in the field of human-AI combinations does not indicate consensus in terms of its effectiveness. For example, using a meta-analysis of studies published between 2020 and 2023, Vaccaro, Almaatouq, and Malone (\citeproc{ref-vaccaro2024combinations}{2024}) finds that, on average, human-AI combinations performed significantly worse than the best of humans or AI alone. On the other hand, Li et al. (\citeproc{ref-li2024explanatory}{2024}) uses a sample of 200 Chinese students to show significant correlations between human--computer interactions, perceived usefulness and students' skills. Understanding these associations can help institutions design better AI courses and more effective integration of AI in education.

This study advances this literature on human-AI collaboration in educational contexts by examining the role of step-by-step reasoning and the impact of different modes of AI interaction. Additionally, it contributes to ongoing research exploring the factors that influence students' engagement with AI. Gaining insights into the value these combinations bring is crucial for developing practical guidelines for schools and educational institutions.

Results of our investigation also contribute to the current debate on the pedagogical principles for designing AI-driven systems and educational opportunities, the so-called \emph{design for learning} (\citeproc{ref-gavsevic2023empowering}{Gašević, Siemens, and Sadiq 2023}). It underscores the importance of using a multi-disciplinary approach that combines traditional pedagogical insights with principles from human-computer interactions. Consequently, our work adds to the growing efforts to tackle critical challenges in integrating AI into education, helping to automate and scale up tasks like providing feedback, grading, and making learning recommendations (\citeproc{ref-dai2023can}{Dai et al. 2023}; \citeproc{ref-yan2024practical}{Yan, Sha, et al. 2024}; \citeproc{ref-joksimovic2023opportunities}{Joksimovic et al. 2023}).
